\HeaderA{dirs}{Get subdirectories}{dirs}
\keyword{IO}{dirs}
\begin{Description}\relax
If you want only subfolders and no files, use \code{dirs}.  
With \code{recursive} = FALSE, \code{\LinkA{dir}{dir}} returns both folders
and files.  With \code{recursive} = TRUE, it returns only files.
\end{Description}
\begin{Usage}
\begin{verbatim}
dirs(path='.', pattern=NULL, exclude=NULL, all.files=FALSE,
     full.names=FALSE, recursive=FALSE, ignore.case=FALSE) 
\end{verbatim}
\end{Usage}
\begin{Arguments}
\begin{ldescription}
\item[\code{path, all.files, full.names, recursive, ignore.case}] as for \code{\LinkA{dir}{dir}}

\item[\code{pattern, exclude}] optional regular expressions of filenames to include or exclude,
respectively.  

\end{ldescription}
\end{Arguments}
\begin{Details}\relax
1.  mainDir <- dir(...)  without recurse 

2.  Use \code{\LinkA{file.info}{file.info}} to restrict mainDir to only
directories.

3.  If !recursive, return the restricted mainDir.  Else, if
length(mainDir) > 0, create dirList to hold the results of the
recursion and call \code{dirs} for each component of mainDir.  Then
\code{\LinkA{unlist}{unlist}} and return the result.
\end{Details}
\begin{Value}
A character vector of the desired subdirectories.
\end{Value}
\begin{Author}\relax
Spencer Graves
\end{Author}
\begin{SeeAlso}\relax
\code{\LinkA{dir}{dir}},
\code{\LinkA{file.info}{file.info}}
\code{\LinkA{package.dir}{package.dir}}
\end{SeeAlso}
\begin{Examples}
\begin{ExampleCode}
path2fdaM <- system.file('Matlab/fdaM', package='fda')
dirs(path2fdaM)
dirs(path2fdaM, full.names=TRUE)
dirs(path2fdaM, recursive=TRUE)
dirs(path2fdaM, exclude='^@|^private$', recursive=TRUE)

# Directories to add to Matlab path
# for R.matlab and fda
R.matExt <- system.file('externals', package='R.matlab')
fdaM <- dirs(path2fdaM, exclude='^@|^private$', full.names=TRUE,
              recursive=TRUE)  
add2MatlabPath <- c(R.matExt, path2fdaM, fdaM) 

\end{ExampleCode}
\end{Examples}

