\HeaderA{lines.fd}{Add Lines from Functional Data to a Plot}{lines.fd}
\aliasA{lines.fdSmooth}{lines.fd}{lines.fdSmooth}
\keyword{smooth}{lines.fd}
\begin{Description}\relax
Lines defined by functional observations are added to an existing plot.
\end{Description}
\begin{Usage}
\begin{verbatim}
## S3 method for class 'fd':
lines(x, Lfdobj=int2Lfd(0), nx=201, ...)
## S3 method for class 'fdSmooth':
lines(x, Lfdobj=int2Lfd(0), nx=201, ...)
\end{verbatim}
\end{Usage}
\begin{Arguments}
\begin{ldescription}
\item[\code{x}] a univariate functional data object to be evaluated at \code{nx}
points over \code{xlim} and added as a line to an existing plot.  

\item[\code{Lfdobj}] either a nonnegative integer or a linear differential operator
object.  If present, the derivative or the value of applying the
operator is evaluated rather than the functions themselves.

\item[\code{nx}] Number of points within \code{xlim} at which to evaluate \code{x}
for plotting.  

\item[\code{... }] additional arguments such as axis titles and so forth that can be
used in plotting programs called by \code{lines.fd} or
\code{lines.fdSmooth}.  

\end{ldescription}
\end{Arguments}
\begin{Section}{Side Effects}
Lines added to an existing plot.
\end{Section}
\begin{SeeAlso}\relax
\code{\LinkA{plot.fd}{plot.fd}}, 
\code{\LinkA{plotfit.fd}{plotfit.fd}}
\end{SeeAlso}
\begin{Examples}
\begin{ExampleCode}
##
## plot a fit with 3 levels of smoothing
##
x <- seq(-1,1,0.02)
y <- x + 3*exp(-6*x^2) + sin(1:101)/2
# sin not rnorm to make it easier to compare
# results across platforms 

result4. <- smooth.basisPar(argvals=x, y=y, lambda=1)
result4.4 <- smooth.basisPar(argvals=x, y=y, lambda=1e-4)
result4.0 <- smooth.basisPar(x, y, lambda=0)

plot(x, y)
lines(result4.)
lines(result4.4, col='green')
lines.fdSmooth(result4.0, col='red') 

plot(x, y, xlim=c(0.5, 1))
lines.fdSmooth(result4.)
lines.fdSmooth(result4.4, col='green')
lines.fdSmooth(result4.0, col='red')  
lines.fdSmooth(result4.0, col='red', nx=101)
# no visible difference from the default?  

lines.fdSmooth(result4.0, col='orange', nx=31)
# Clear difference, especially between 0.95 and 1  

\end{ExampleCode}
\end{Examples}

