\HeaderA{plot.pca.fd}{Plot Functional Principal Components}{plot.pca.fd}
\keyword{smooth}{plot.pca.fd}
\begin{Description}\relax
Display the types of variation across a sample of functions.  Label
with the eigenvalues that indicate the relative importance of each
mode of variation.
\end{Description}
\begin{Usage}
\begin{verbatim}
plot.pca.fd(x, nx = 128, pointplot = TRUE, harm = 0,
                        expand = 0, cycle = FALSE, ...)
\end{verbatim}
\end{Usage}
\begin{Arguments}
\begin{ldescription}
\item[\code{x}] a functional data object.

\item[\code{nx}] Number of points to plot or vector (if length > 1) to use as
\code{evalarg} in evaluating and plotting the functional principal
components. 

\item[\code{pointplot}] logical:  If TRUE, the harmonics / principal components are plotted
as '+' and '-'.   Otherwise lines are used.

\item[\code{harm}] Harmonics / principal components to plot.  If 0, plot all.

If length(harm) > sum(par("mfrow")), the user advised, "Waiting to
confirm page change..." and / or 'Click or hit ENTER for next page'
for each page after the first.  

\item[\code{expand}] nonnegative real:  If expand == 0 then effect of +/- 2 standard
deviations of each pc are given otherwise the factor expand is
used.  

\item[\code{cycle}] logical:  If cycle=TRUE and there are 2 variables then a cycle plot
will be drawn If the number of variables is anything else, cycle
will be ignored. 

\item[\code{...}] other arguments for 'plot'.  

\end{ldescription}
\end{Arguments}
\begin{Details}\relax
Produces one plot for each principal component / harmonic to be
plotted.
\end{Details}
\begin{Value}
invisible(NULL)
\end{Value}
\begin{SeeAlso}\relax
\code{\LinkA{cca.fd}{cca.fd}}, 
\code{\LinkA{pda.fd}{pda.fd}}
\code{\LinkA{plot.pca.fd}{plot.pca.fd}}
\end{SeeAlso}
\begin{Examples}
\begin{ExampleCode}

#  carry out a PCA of temperature
#  penalize harmonic acceleration, use varimax rotation

daybasis65 <- create.fourier.basis(c(0, 365), nbasis=65, period=365)

harmaccelLfd <- vec2Lfd(c(0,(2*pi/365)^2,0), c(0, 365))
harmfdPar     <- fdPar(daybasis65, harmaccelLfd, lambda=1e5)
daytempfd <- data2fd(CanadianWeather$dailyAv[,,"Temperature.C"],
      day.5, daybasis65, argnames=list("Day", "Station", "Deg C"))

daytemppcaobj <- pca.fd(daytempfd, nharm=4, harmfdPar)
#  plot harmonics, asking before each new page after the first:  
plot.pca.fd(daytemppcaobj)

# plot 4 on 1 page
op <- par(mfrow=c(2,2))
plot.pca.fd(daytemppcaobj, cex.main=0.9)
par(op)

\end{ExampleCode}
\end{Examples}

