\HeaderA{tperm.fd}{Permutation t-test for two groups of functional data objects.}{tperm.fd}
\keyword{smooth}{tperm.fd}
\begin{Description}\relax
tperm.fd creates a null distribution for a test of no difference between two
groups of functional data objects.
\end{Description}
\begin{Usage}
\begin{verbatim}
 tperm.fd(x1fd, x2fd, nperm=200, q=0.05, argvals=NULL, plotres=TRUE,...)
\end{verbatim}
\end{Usage}
\begin{Arguments}
\begin{ldescription}
\item[\code{x1fd}] a functional data object giving the first group of functional observations.

\item[\code{x2fd}] a functional data object giving the second group of functional
observations.

\item[\code{nperm}] number of permutations to use in creating the null distribution.

\item[\code{argvals}] If \code{yfdPar} is a \code{fd} object, the points at which to evaluate
the point-wise F-statistic.

\item[\code{q}] Critical upper-tail quantile of the null distribution to compare to the observed F-statistic.

\item[\code{plotres}] Argument to plot a visual display of the null distribution displaying the
\code{1-q}th quantile and observed F-statistic.

\item[\code{...}] Additional plotting arguments that can be used with \code{plot}.

\end{ldescription}
\end{Arguments}
\begin{Details}\relax
The usual t-statistic is calculated pointwise and the test based on the
maximal value. If \code{argvals} is not specified,
it defaults to 101 equally-spaced points on the range of \code{yfdPar}.
\end{Details}
\begin{Value}
A list with components
\begin{ldescription}
\item[\code{pval}] the observed p-value of the permutation test.
\item[\code{qval}] the \code{q}th quantile of the null distribution.
\item[\code{Tobs}] the observed maximal t-statistic.
\item[\code{Tnull}] a vector of length \code{nperm} giving the observed values of the
permutation distribution.

\item[\code{Tvals}] the pointwise values of the observed t-statistic.
\item[\code{Tnullvals}] the pointwise values of of the permutation observations.
\item[\code{pvals.pts}] pointwise p-values of the t-statistic.
\item[\code{qvals.pts}] pointwise \code{q}th quantiles of the null distribution
\item[\code{argvals}] argument values for evaluating the F-statistic if \code{yfdPar}is
a functional data object.

\end{ldescription}

normal-bracket50bracket-normal
\end{Value}
\begin{Section}{Side Effects}
a plot of the functional observations
\end{Section}
\begin{Source}\relax
Ramsay, James O., and Silverman, Bernard W. (2006), \emph{Functional
Data Analysis, 2nd ed.}, Springer, New York.
\end{Source}
\begin{SeeAlso}\relax
\code{\LinkA{fRegress}{fRegress}}
\code{\LinkA{Fstat.fd}{Fstat.fd}}
\end{SeeAlso}
\begin{Examples}
\begin{ExampleCode} 
# This tests the difference between boys and girls heights in the Berkeley
# growth data.

# First set up a basis system to hold the smooths

knots  <- growth$age
norder <- 6
nbasis <- length(knots) + norder - 2
hgtbasis <- create.bspline.basis(range(knots), nbasis, norder, knots)

# Now smooth with a fourth-derivative penalty and a very small smoothing
# parameter

Lfdobj <- 4
lambda <- 1e-2
growfdPar <- fdPar(hgtbasis, Lfdobj, lambda)

hgtmfd <- smooth.basis(growth$age, growth$hgtm, growfdPar)$fd
hgtffd <- smooth.basis(growth$age, growth$hgtf, growfdPar)$fd

# Call tperm.fd

tres <- tperm.fd(hgtmfd,hgtffd)
\end{ExampleCode}
\end{Examples}

